Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-01-28 man/ms.Rmd
update changes intro

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017]. Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for at least the last three decades [@Spinoni2017b]. Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013], specially in the Mediterranean area [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017]. In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014]. The functional response of vegetation to drought has been analyzed at different scales, from global [*e.g*. @Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].

Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification such as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016]. `REVIEW` For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.

The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution `REVIEW`[@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015]. The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth. Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Zribi2016].

Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened. We are also interested in assessing resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades. We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

>>>>

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Dai2011; @Ummenhofer2017]. Despite drought is considered a feature of Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events has been recorded in the last decades [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported [@Spinoni2017b]. This is especially important, given the climate change projections for the Mediterranean area that forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017]. The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global [*e.g.*, @VicenteSerrano2013; @Norman2016] and at local scales [see @MartinezVilalta2016 for a revision].

Ecosystem responses to drought are influenced by other drivers of environmental changes [@Fischer2006; @Oliver2014; @Penuelas2017]. In a global-change context, the impacts of drought on vegetation need to be evaluate considering the interactions with other drivers of change [@Clavero2011; @DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change is a key driver [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution
[*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018]. Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. The assessment of resilience --*i.e.* the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015]-- to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of the drought for drier biomes [@VicenteSerrano2013]. For Mediterranean, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].

Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].

The combination of both approaches, namely remote sensing and dendroecology, has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]

Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened. We also assess the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.   
Our hypothesis is that these relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of Q. pyrenaica forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
